---
title: "Basic statistics from the occupancy interview data"
output: html_notebook
---

## Basic statistics from the occupancy interview data

let's start with loading our data

```{r}
#import packages
library(readr)
library(dplyr)
library(ggplot2)

#load covariate and occurence data
covariates <- read_tsv("data/Occupancy Data Collection Sheet - covariates master sheet.tsv")
# merged species occurence records for 2024-2025
occ_cur <- read_tsv("data/Occupancy Data Collection Sheet - hornbill detections current.tsv")

# merged species occurence records for 2004-2005
occ_pas <- read_tsv("data/Occupancy Data Collection Sheet - hornbill detections past.tsv")


```

## Now we will see some basic stats

```{r}
grids <- unique(covariates$grid.id)
# so we sampled 96 grids in total
#now make a histogram of number of replicates in these grids

# add a replicates column per grid.id
covariates <- covariates %>%
  add_count(grid.id, name = "replicates")


# filter to keep only the interviews with current detections
covariates_cur <- covariates %>% filter(proportion.grid.24 != "na")

# Count replicates per grid.id
replicate_counts <- covariates_cur %>%
  count(grid.id)
# Histogram of those counts
ggplot(replicate_counts, aes(x = n)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
   geom_text(
    stat = "bin", binwidth = 1,
    aes(label = after_stat(count)),
    vjust = -0.5, size = 3.5
  ) +
  labs(
    title = "Distribution of Replicates per Grid ID",
    x = "Number of Replicates",
    y = "Number of Grid IDs"
  ) +
  theme_minimal()
```

Thus we have 95 grids that have at least one replicate and 89 with at least three replicates for current occupancy.

Now for the historic data (2004-2005)

```{r}
# filter to keep only the interviews with historical detections
covariates_pas <- covariates %>% filter(proportion.grid.04 != "na")

# Count replicates per grid.id
replicate_counts <- covariates_pas %>%
  count(grid.id)
# Histogram of those counts
ggplot(replicate_counts, aes(x = n)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
   geom_text(
    stat = "bin", binwidth = 1,
    aes(label = after_stat(count)),
    vjust = -0.5, size = 3.5
  ) +
  labs(
    title = "Distribution of Replicates per Grid ID",
    x = "Number of Replicates",
    y = "Number of Grid IDs"
  ) +
  theme_minimal()
```

Thus we have 94 grids that have at least one replicate and 82 with at least three replicates for historic occupancy.

**So just to include 5 and 12 more grids in the** current and historic distribution we should not use the variable sampling area strategy.

## Let's make some maps

We will now switch make a couple of figures showing the spatial extent of our data.

```{r}
setwd("C:/DATA/NCF/Occupancy modelling")

# load the grids shape file 
library(sf)
grids <- st_read("")
```