---
title: "Basic statistics from the occupancy interview data"
output: html_notebook
---

## Basic statistics from the occupancy interview data

let's start with loading our data

```{r}
#import packages
library(readr)
library(dplyr)
library(ggplot2)
library(sf)
library(stringr)


#load covariate and occurence data
covariates <- read_tsv("data/raw/covariates-cur-his-clean.txt")
# merged species occurence records for 2024-2025
occ_cur <- read_tsv("data/Occupancy Data Collection Sheet - hornbill detections current.tsv")

# merged species occurence records for 2004-2005
occ_pas <- read_tsv("data/Occupancy Data Collection Sheet - hornbill detections past.tsv")


```

## Now we will see some basic stats

```{r}
grids <- unique(covariates$grid.id)
# so we sampled 96 grids in total
#now make a histogram of number of replicates in these grids

# add a replicates column per grid.id
covariates <- covariates %>%
  add_count(grid.id, name = "replicates")


# filter to keep only the interviews with current detections
covariates_cur <- covariates %>% filter(proportion.grid.24 != "na")

# Count replicates per grid.id
replicate_counts <- covariates_cur %>%
  count(grid.id)
# Histogram of those counts
ggplot(replicate_counts, aes(x = n)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
   geom_text(
    stat = "bin", binwidth = 1,
    aes(label = after_stat(count)),
    vjust = -0.5, size = 3.5
  ) +
  labs(
    title = "Distribution of Replicates per Grid ID",
    x = "Number of Replicates",
    y = "Number of Grid IDs"
  ) +
  theme_minimal()
```

Thus we have 95 grids that have at least one replicate and 89 with at
least three replicates for current occupancy.

Now for the historic data (2004-2005)

```{r}
# filter to keep only the interviews with historical detections
covariates_pas <- covariates %>% filter(proportion.grid.04 != "na")

# Count replicates per grid.id
replicate_counts <- covariates_pas %>%
  count(grid.id)
# Histogram of those counts
ggplot(replicate_counts, aes(x = n)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
   geom_text(
    stat = "bin", binwidth = 1,
    aes(label = after_stat(count)),
    vjust = -0.5, size = 3.5
  ) +
  labs(
    title = "Distribution of Replicates per Grid ID",
    x = "Number of Replicates",
    y = "Number of Grid IDs"
  ) +
  theme_minimal()
```

Thus we have 94 grids that have at least one replicate and 82 with at
least three replicates for historic occupancy.

**So just to include 5 and 12 more grids in the** current and historic
distribution we should not use the variable sampling area strategy.

## Let's make some maps

We will now make a couple of figures showing the spatial extent of our
data.

```{r}
# load the grids shape file 
# Unzip the KMZ to a temporary directory
kmz_path <- "../spatial-data/Grids.kmz"
unzip(kmz_path, exdir = tempdir())

# Find the extracted KML file
kml_path <- list.files(tempdir(), pattern = "\\.kml$", 
                        full.names = TRUE, recursive = TRUE)

# Read it
grids <- st_read(kml_path)
covariates <- read_tsv("data/raw/covariates-cur-his-clean.txt")

## Now that we understand the structure of the kml we will proceed to extract it as a dataframe

#structuring ht grids_sf file for filtering
#make a function to extract fid, mean_elev, and elev_band
extract_field <- function(html, field_name) {
  pattern <- paste0("(?<=<b>", field_name, "</b></small></td><td[^>]{0,50}>)[^<]+")
  str_extract(html, pattern)
}

## now we also edit the sf object to add these columns separately

grids_sf <- grids %>%
  dplyr::mutate(
    fid       = as.integer(extract_field(Description, "fid")),
    mean_elev = as.numeric(extract_field(Description, "mean_elev")),
    elev_band = trimws(extract_field(Description, "elev_band"))
  )

head(grids_sf[, c("fid", "mean_elev", "elev_band")])

covariates <- covariates %>%
  dplyr::left_join(
    grids_sf %>% dplyr::select(fid, mean_elev, elev_band),
    by = c("grid.id" = "fid")
  )
#save this updated covariates
write_tsv(covariates, "data/processed/covariates-full-elevation.txt")
#keep only those grids in grids_sf which have > 3 replicates
covariates <- covariates %>% filter(replicates > 2)
write_tsv(covariates, "data/processed/covariates-full-elevation-filtered.txt")

class(grids_sf$fid)
class(covariates$grid.id)
# Check for NAs
sum(is.na(grids_sf$fid))
sum(is.na(covariates$grid.id))

# Convert both to integer before filtering
grids_df_fil <- grids_sf %>%
  dplyr::filter(as.integer(fid) %in% as.integer(covariates$grid.id))
#remove duplicates
grids_df_fil <- grids_df_fil %>%
  dplyr::distinct(fid, .keep_all = TRUE)
#save this as kml
st_write(grids_df_fil, "data/processed/grids_clean_data_complete.kml", driver = "KML", delete_dsn = TRUE)

```

### Distribution of sampling along the elevational gradient

Now we will make a histogram of number of grids sampled at least thrice
for current and historic data

**Let** **us do it for current data first**

```{r}
# we will use the grids_sf object for this

```
