---
title: "Determining Naive Occupancy"
output: html_notebook
---

### 2 We will now calculate the Naive occupancy

Naive Occupancy is nothing but a ratio of sites where the focus species
was detected and the total number of sites surveyed. For us sampled
sites are grids surveyed more than three times.

```{r}
# Load libraries
library(readr)
library(dplyr)
library(ggplot2)
library(sf)
library(stringr)
library(tidyr)
```

#### 2.1 load and prepare the data

```{r}
# load our current detections data for all species
det_cur <- read_tsv("data/raw/all-hornbill-detections-current.txt")
# restify the concatenated detection history column
det_cur <- det_cur %>% select(-interview.1interview.2interview.3interview4)
det_cur <- det_cur %>%
  dplyr::mutate(
    det.hist = paste0(
      ifelse(is.na(interview.1), "", as.character(interview.1)),
      ifelse(is.na(interview.2), "", as.character(interview.2)),
      ifelse(is.na(interview.3), "", as.character(interview.3)),
      ifelse(is.na(interview4),  "", as.character(interview4))
    )
  )
det_cur_fil <- det_cur %>% filter(interview.3 !="NA") 
# Separate species and filter grids with < 3 replicates
det_cur_gh <- det_cur %>% filter(det_cur$species == "GH", interview.3 !="NA")
det_cur_wh <- det_cur %>% filter(det_cur$species == "WH", interview.3 !="NA")
det_cur_oph <- det_cur %>% filter(det_cur$species == "OPH", interview.3 !="NA")
det_cur_rnh <- det_cur %>% filter(det_cur$species == "RNH", interview.3 !="NA")

```

2.2 Now lets get the Naive occupancy!

```{r}
# we will count the number of grids with at least one detection using out det.hit column

# for great hornbill
#grids with positive detections
det_cur_gh_pos <- filter(det_cur_gh, det_cur_gh$interview.1 == 2 | det_cur_gh$interview.2 == 2 | det_cur_gh$interview.3 == 2 | det_cur_gh$interview4 == 2)
no_gh <- nrow(det_cur_gh_pos)/nrow(det_cur_gh)

#for wreathed hornbill
det_cur_wh_pos <- filter(det_cur_wh, det_cur_wh$interview.1 == 2 | det_cur_wh$interview.2 == 2 | det_cur_wh$interview.3 == 2 | det_cur_wh$interview4 == 2)
no_wh <- nrow(det_cur_wh_pos)/nrow(det_cur_wh)

#for oph
det_cur_pos <- filter(det_cur_oph, det_cur_oph$interview.1 == 2 | det_cur_oph$interview.2 == 2 | det_cur_oph$interview.3 == 2 | det_cur_oph$interview4 == 2)
no_oph <- nrow(det_cur_pos)/93


#for oph
det_cur_pos <- filter(det_cur_rnh, det_cur_rnh$interview.1 == 2 | det_cur_rnh$interview.2 == 2 | det_cur_rnh$interview.3 == 2 | det_cur_rnh$interview4 == 2)
no_rnh <- nrow(det_cur_pos)/93

print(paste("Naive occupancy of GH is",no_gh))
print(paste("Naive occupancy of WH is",no_wh))
print(paste("Naive occupancy of OPH is",no_oph))
print(paste("Naive occupancy of RNH is",no_rnh))
```

#### 2.3 Now repeat the same for historic data

```{r}
# load our current detections data for all species
det_cur <- read_tsv("data/raw/all-hornbill-detections-historic.tsv")
# restify the concatenated detection history column
det_cur <- det_cur %>% select(-detection.history)
det_cur <- det_cur %>%
  dplyr::mutate(
    det.hist = paste0(
      ifelse(is.na(interview.1), "", as.character(interview.1)),
      ifelse(is.na(interview.2), "", as.character(interview.2)),
      ifelse(is.na(interview.3), "", as.character(interview.3)),
      ifelse(is.na(interview4),  "", as.character(interview4))
    )
  )

# Separate species and filter grids with < 3 replicates
det_cur_gh <- det_cur %>% filter(det_cur$species == "GH", interview.3 !="NA")
det_cur_wh <- det_cur %>% filter(det_cur$species == "WH", interview.3 !="NA")
det_cur_oph <- det_cur %>% filter(det_cur$species == "OPH", interview.3 !="NA")
det_cur_rnh <- det_cur %>% filter(det_cur$species == "RNH", interview.3 !="NA")

# we will count the number of grids with at least one detection using out det.hit column

# for great hornbill
#grids with positive detections
det_cur_gh_pos <- filter(det_cur_gh, det_cur_gh$interview.1 == 2 | det_cur_gh$interview.2 == 2 | det_cur_gh$interview.3 == 2 | det_cur_gh$interview4 == 2)
no_gh <- nrow(det_cur_gh_pos)/nrow(det_cur_gh)

#for wreathed hornbill
det_cur_wh_pos <- filter(det_cur_wh, det_cur_wh$interview.1 == 2 | det_cur_wh$interview.2 == 2 | det_cur_wh$interview.3 == 2 | det_cur_wh$interview4 == 2)
no_wh <- nrow(det_cur_wh_pos)/nrow(det_cur_wh)

#for oph
det_cur_pos <- filter(det_cur_oph, det_cur_oph$interview.1 == 2 | det_cur_oph$interview.2 == 2 | det_cur_oph$interview.3 == 2 | det_cur_oph$interview4 == 2)
no_oph <- nrow(det_cur_pos)/86


#for oph
det_cur_pos <- filter(det_cur_rnh, det_cur_rnh$interview.1 == 2 | det_cur_rnh$interview.2 == 2 | det_cur_rnh$interview.3 == 2 | det_cur_rnh$interview4 == 2)
no_rnh <- nrow(det_cur_pos)/86

print(paste("Naive occupancy of GH is",no_gh))
print(paste("Naive occupancy of WH is",no_wh))
print(paste("Naive occupancy of OPH is",no_oph))
print(paste("Naive occupancy of RNH is",no_rnh))

# save both the filtered datasets
det_his_fil <- det_his %>% filter(interview.3 !="NA") 

write_tsv(det_his_fil, "data/processed/all-hornbill-historic-detections-filtered.txt")
write_tsv(det_cur_fil, "data/processed/all-hornbill-current-detections-filtered.txt")
write_tsv(det_his, "data/raw/all-hornbill-detections-historic.txt")
```

3 Technically we are done for now however we will do some more
restructuring that can be useful in some later analyses

```{r}
# Replace useless highno. columns by a single one
```
