---
title: "Cleaning the covariates"
author: "Ali Khan Faizee"
date: "2026-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Road length

Let's start with preparing the road length per grid covariates. We have
already calculated the length of roads in the grids with QGIS, now will
add them to the grid data

```{r}
# load libraries and data
library(sf)
library(dplyr)
library(stringr)
road_length <- read.csv("spatial-data/covariate-calcs/road-lengths.csv")
head(road_length)
# Check results
sum(road_length$road_length == 0)   # how many grids have no roads

# Join to covariates
covariates <- read.delim("data/processed/covariates-full-elevation-filtered.txt", sep = "\t")

covariates_road <- covariates %>%
  dplyr::left_join(
    road_length %>% select(fid, road_length),
    by = c("grid.id" = "fid")
  )
## let us loose the useless columns in the covariates 

covariates_road <- covariates_road %>% select(-mean_elev.x, -elev_band.x, -geometry.x)
covariates_road <- covariates_road %>% mutate(mean_elev = mean_elev.y, elev_band = elev_band.y, geometry = geometry.y) %>% select(-mean_elev.y, -elev_band.y, -geometry.y)
write.table(covariates_road, "data/processed/covariates-full-elevation-roads.txt", sep = "\t")
```

I realized that the grids are not all square shaped so I will replace
the geometry with an un adulterated grid set

```{r}
data_grids_clean <- st_read("spatial-data/grids_clean_data_complete.kml")
data_grids <- st_read("spatial-data/grids_with_data.kml")
naive_grids <- st_read("spatial-data/naive_grids.shp")


#make a function to extract fid, mean_elev, and elev_band
extract_field <- function(html, field_name) {
  pattern <- paste0("(?<=<b>", field_name, "</b></small></td><td[^>]{0,50}>)[^<]+")
  str_extract(html, pattern)
}

## now we also edit the sf object to add these columns separately
data_grids_clean <- data_grids_clean %>%
  dplyr::mutate(
    fid       = as.integer(extract_field(Description, "fid")),
    mean_elev = as.numeric(extract_field(Description, "mean_elev")),
    elev_band = trimws(extract_field(Description, "elev_band"))
  )
#save this
st_write(data_grids_clean, "spatial-data/grids-with-data.gpkg")

#now its turn for all the grids
data_grids <- data_grids %>%
  dplyr::distinct(fid, .keep_all = TRUE)
#remove duplicates from the grids with data
data_grids <- data_grids %>%
  dplyr::mutate(
    fid       = as.integer(extract_field(Description, "fid")),
    mean_elev = as.numeric(extract_field(Description, "mean_elev")),
    elev_band = trimws(extract_field(Description, "elev_band"))
  )
##then replace the geometry according to fid
naive_grids <- naive_grids %>% st_transform(crs = 4326) %>% filter(fid %in% data_grids$fid) %>% dplyr::mutate(fid = as.integer(fid))

st_write(naive_grids, "final-grids-with-data.gpkg")

## correcting sampled grids
naive_grids <- st_read("spatial-data/final-grids-with-data.shp")
sampled_grids <- naive_grids %>% st_transform(crs = 4326) %>% filter(fid %in% data_grids_clean$fid) %>% dplyr::mutate(fid = as.integer(fid))
st_write(sampled_grids, "spatial-data/sampled-grids.gpkg")
```

### Now let us do canopy height

```{r}
library(terra)

# List all tif files in folder
tiles <- list.files("spatial-data/canopy-height-2020/", pattern = "\\.tif$", full.names = TRUE)

# Now we will calculate mean canopy height in our grids

# Match CRS
grids <- st_read("spatial-data/covariate-calcs/final-grids-with-data.shp")
raster <- terra::rast("spatial-data/canopy-height-2020/ETH_GlobalCanopyHeight_10m_2020_N24E090_Map.tif")
grids <- st_transform(grids, crs(raster))

# Extract mean per grid polygon
vrt_raster <- vrt(tiles)

# Use it directly for extraction without saving
values_canopy_height <- terra::extract(vrt_raster, vect(grids), fun = mean, na.rm = TRUE)

grids <- grids %>% mutate(fid = as.integer(fid))
st_write(grids, "final-grids-with-data-canopy.gpkg")
write.csv(grids, "final-grids-with-data-canopy.csv")
```

Add these to the covariates table so that we can get rid of the
subsidairy variables

```{r}
covariates <- read.delim("data/processed/covariates-full-elevation-roads.txt")

covariates <- covariates %>%
  dplyr::left_join(
    grids %>% dplyr::select(fid, canopy_height),  # replace with actual column name
    by = c("grid.id" = "fid")
  ) %>% select(-geometry.x, -canopy_height.x) %>% rename(geometry = geometry.y, canopy_height=canopy_height.y)

## add slope
slope <- st_read("final-grids-with-data-slope.shp")

covariates <- covariates %>%
  dplyr::left_join(
    slope %>% dplyr::select(fid, slope),
    by = c("grid.id" = "fid")
  ) 
covariates <- covariates %>% select(-geometry.x) %>% rename(geometry = geometry.y)

### repeat for aspect

aspect_rast <- rast("spatial-data/covariate-calcs/dem_ap_srtm_aspect.tif")

grids <- st_read("spatial-data/covariate-calcs/final-grids-with-data.shp")

grids_aspect <- terra::extract(aspect_rast, vect(grids), fun = mean, na.rm = TRUE)
grids$aspect <- grids_aspect$dem_ap_srtm_aspect
covariates <- covariates %>%
  dplyr::left_join(
    grids %>% dplyr::select(fid, aspect),
    by = c("grid.id" = "fid")
  ) 
covariates <- covariates %>% select(-geometry.x) %>% rename(geometry = geometry.y)

grids$canopy_height <- values_canopy_height$spat_57cc573f7c9c_22476_zxzxx9jyyx6UnnI
grids$slope <- slope$slope

write.csv(grids, "spatial-data/covariate-calcs/grids-with-aspect-canopy-height-slope.csv")
write_tsv(covariates, "data/processed/covariates-full-elevation-roads-aspect-canopy-height-slope.txt")
```

### Area under various classes

```{r}
library(sf)
library(dplyr)

# Load LULC shapefile
lulc <- st_read("spatial-data/area-under/LULC/NRSC_LULC.shp")
grids <- st_read("spatial-data/covariate-calcs/final-grids-with-data.shp")   # or .gpkg

# Check column names — find which column has the class names
names(lulc)
head(lulc)

# Check unique class names (exact spelling matters)
  unique(lulc$LULC_Class)   # replace 'class' with your actual column name
  
st_crs(lulc)
st_crs(grids)

# Project both to UTM 46N for area in metres
lulc  <- st_transform(lulc,  crs = 32646)
grids <- st_transform(grids, crs = 32646)

# Fix invalid geometries
lulc  <- st_make_valid(lulc)
grids <- st_make_valid(grids)

classes_of_interest <- c(
  "Shifting Cultivation",
  "Scrub Forest",
  "Evergreen Forest\r\nEvergreen Forest",
  "Deciduous Forest",
  "Alpine/Sub Alpine",
  "Agricuture"
)

# Filter — adjust names to exactly match unique(lulc$class) output
lulc_filtered <- lulc %>%
  dplyr::filter(LULC_Class %in% classes_of_interest)

lulc_in_grids <- st_intersection(lulc_filtered, grids)

lulc_area <- lulc_in_grids %>%
  dplyr::mutate(area_m2 = as.numeric(st_area(geometry))) %>%
  dplyr::group_by(fid, LULC_Class) %>%
  dplyr::summarise(area_m2 = sum(area_m2, na.rm = TRUE)) %>%
  st_drop_geometry()

head(lulc_area)

library(tidyr)
library(janitor)

lulc_wide <- lulc_area %>%
  tidyr::pivot_wider(
    names_from  = LULC_Class,
    values_from = area_m2,
    values_fill = 0        # grids with no coverage of a class get 0
  )

# Clean column names
names(lulc_wide) <- janitor::make_clean_names(names(lulc_wide))

write.table(lulc_wide, "area-under-lulc-grids.txt", sep = "\t")

### Now add them to the covariates table 
covariates <- read.delim("data/processed/covariates-full-elevation-roads-aspect-canopy-height-slope.txt")

covariates_area <- covariates %>%
  dplyr::left_join(
    lulc_wide,
    by = c("grid.id" = "fid")
  )


covariates_area$evergreen_deciduous_forest <- covariates_area$evergreen_forest_evergreen_forest + covariates_area$deciduous_forest

covariates_area <- select(covariates_area, -evergreen_forest_evergreen_forest, -deciduous_forest)

write.table(covariates_area, "spatial-data/area-under/covariates-full-elevation-roads-aspect-canopy-height-slope-area-under.txt", sep = "\t")
```

**Now we need to do the same for protected areas**

```{r}

# Load PAs shapefile
pas <- st_read("spatial-data/area-under/PAs/TR-PA-corridor-NTCA.shp")
grids <- st_read("spatial-data/covariate-calcs/final-grids-with-data.shp")   # or .gpkg

# Check column names — find which column has the class names
names(pas)
head(pas)

# Check unique class names (exact spelling matters)
  unique(lulc$LULC_Class)   # replace 'class' with your actual column name
  
st_crs(pas)
st_crs(grids)

# Project both to UTM 46N for area in metres
pas  <- st_transform(pas,  crs = 32646)
grids <- st_transform(grids, crs = 32646)

# Fix invalid geometries
pas  <- st_make_valid(pas)
grids <- st_make_valid(grids)


pas_in_grids <- st_intersection(pas, grids)

pas_area <- pas_in_grids %>%
  dplyr::mutate(area_m2 = as.numeric(st_area(geometry))) %>%
  dplyr::group_by(fid) %>%
  dplyr::summarise(area_m2 = sum(area_m2)) %>%
  st_drop_geometry()

head(pas_area)

# Clean column names
pas_area <- pas_area %>%
  dplyr::rename(area_pa = area_m2)

### Now add them to the covariates table 
covariates <- read.delim("spatial-data/area-under/covariates-full-elevation-roads-aspect-canopy-height-slope-area-under.txt")

covariates_pa_area <- covariates %>%
  dplyr::left_join(
    pas_area,
    by = c("grid.id" = "fid")
  )


write.table(covariates_pa_area, "spatial-data/area-under/covariates-full-elevation-roads-aspect-canopy-height-slope-area-under-pa.txt", sep = "\t")
```

Now we will add NDVI as a co-variate, which will serve as a proxy for
canopy cover.

```{r}

covariates <- read.delim("spatial-data/area-under/covariates-full-elevation-roads-aspect-canopy-height-slope-area-under-pa.txt")
ndvi_nov <- read.csv("spatial-data/covariate-calcs/all-grids-data-ndvi-nov-mean.csv")
ndvi_nov <- ndvi_nov[, c(1,5)]
covariates <- covariates %>%
  dplyr::left_join(
    ndvi_nov %>% dplyr::select(fid, NDVI_maymean),  # replace with actual column name
    by = c("grid.id" = "fid")
  ) 

write.table(covariates, "spatial-data/area-under/covariates-full-elevation-roads-aspect-canopy-height-slope-area-under-pa-ndvi.txt", sep = "\t")


```

Now add population as another co-variate

```{r}
covariates <- read.delim("spatial-data/area-under/covariates-full-elevation-roads-aspect-canopy-height-slope-area-under-pa-ndvi.txt")
pop_2015 <- read.csv("spatial-data/covariate-calcs/population-2015.csv")
covariates <- covariates %>%
  dplyr::left_join(
    pop_2015 %>% dplyr::select(fid, X_populationmean),  # replace with actual column name
    by = c("grid.id" = "fid")
  ) 

write.table(covariates, "spatial-data/area-under/covariates-full-elevation-roads-aspect-canopy-height-slope-area-under-pa-ndvi-population.txt", sep = "\t")


```
