---
title: "R Notebook"
output: html_notebook
---

Load necessary data

```{r}
library(dplyr)
library(tidyr)
library(unmarked)
library(AICcmodavg)
library(combinat)
```

Modelling for Great Hornbill

```{r}
det.cur <- read.delim("results/gh-det-cur-ready.txt") %>%
  dplyr::mutate(across(3:6, ~ifelse(. == 2, 1, .))) 

site.cov.selected <- read.delim("results/site-covariate-cur-ready.txt") %>% select(-grid.id)
exp_grid <- read.delim("results/exp-current.txt")
visit_grid <- read.delim("results/visit-current.txt")
prop_grid <- read.delim("results/prop-current.txt")

umf<- unmarkedFrameOccu(y = det.cur[,3:6],
                          siteCovs = site.cov.selected,
                          obsCovs = list(experience = exp_grid, 
                                         visit = visit_grid,
                                         proportion = prop_grid)
                         )     # creating unmarked occupancy data frame

summary(umf)

# Null model baseline
fm <- occu(formula = ~ 1 
                     ~ 1,
           data = umf)

summary(fm)
backTransform(fm, type = "state")
## report the Psi values for the null model - GH = 0.3317373	+- 0.0.05195	

#this is the naive occupancy for GH, which matches our previous manual calculation.
backTransform(fm, type = "det")

```

**Now, to add some covariates. First, assuming constant detection:**

### Step 1 — Fit Global Detection Model (Ψ = null)

model selection for *p* while keeping the null model of Ψ constant

```{r}
# Global detection model with null occupancy
#include all the covariates that can affect detection probability
global_det <- occu(~experience + visit + proportion + hunting.24 + shifting_cultivation + evergreen_deciduous_forest + dist.to.grid + canopy_height + road_length + agricuture #detection fx
                   ~1, # occupancy fx (here constant)
                   data = umf) 
summary(global_det)
```

### Step 2 — Simplify Global Detection Model

Exclude covariates where SE \>= \|β\|:

```{r}
# Extract detection coefficients and SEs
det_coefs <- as.data.frame(summary(global_det)$det)
det_coefs

# Flag covariates where SE >= |beta|
det_coefs$exclude <- abs(det_coefs$SE) >= abs(det_coefs$Estimate)
det_coefs

#Keep only covariates where SE < |beta| and refit: remove proportion
simplified_global_det <-occu(~ visit + hunting.24 + shifting_cultivation + evergreen_deciduous_forest + road_length #detection fx
                   ~1, # occupancy fx (here constant)
                   data = umf)
simplified_global_det_summary <- summary(simplified_global_det)
```

### Note: we found that the following are not explanatory covariates for P:

experience, proportion, dist.to.grid, canopy_height, and agriculture

meaning it does not affect the detection probability for Great Hornbill.

#### Step 3 — GOF on Simplified Global Detection Model

```{r}
gof_det <- mb.gof.test(simplified_global_det, nsim = 1000, plot.hist = TRUE)
chat_det <- gof_det$c.hat.est
cat("c-hat for detection:", chat_det, "\n")
#c-hat for detection: 1.067564  
```

### P = 0.293

This is the proportion of simulated statistics that exceeded your
observed statistic. A p-value of **0.408 means no significant lack of
fit** — your model fits the data well. You want p \> 0.05, and 0.408 is
comfortably above that.

### Observed = 34.85

This is your actual test statistic from the data.

**c-hat for detection:**

```         
1.067564 
```

A c-hat close to 1 means **no overdispersion**, so your QAICc and AICc
values will be very similar. Your model assumptions are not violated and
you can proceed

### Step 4 — Build Competing Detection Models

Using only the remaining covariates after simplification:

```{r}
covs <- c("visit", "hunting.24", "shifting_cultivation",      
          "evergreen_deciduous_forest", "road_length")

# Occupancy formula (null)
occ_formula_str <- "~1"

# Generate all combinations
all_combos <- list()

# Null model
all_combos[[1]] <- occu(formula(paste("~1", occ_formula_str)), data = umf)
names(all_combos)[1] <- "null"

# Single covariates (6 models)
for (i in seq_along(covs)) {
  combo_name <- covs[i]
  formula_str <- paste("~", covs[i], occ_formula_str)
  all_combos[[i + 1]] <- occu(formula(formula_str), data = umf)
  names(all_combos)[i + 1] <- combo_name
}

# All combinations of 2+ covariates
counter <- length(all_combos) + 1

for (n in 2:length(covs)) {
  combos_n <- combn(covs, n, simplify = FALSE)
  
  for (combo in combos_n) {
    combo_name <- paste(combo, collapse = "_")
    formula_str <- paste("~", paste(combo, collapse = " + "), occ_formula_str)
    
    tryCatch({
      all_combos[[counter]] <- occu(formula(formula_str), data = umf)
      names(all_combos)[counter] <- combo_name
      counter <- counter + 1
    }, error = function(e) {
      cat("Failed to fit:", combo_name, "\n")
    })
  }
}

# Remove duplicates (just in case)
all_combos <- all_combos[!duplicated(names(all_combos))]
# Remove models with convergence issues
det_model_list <- Filter(function(m) m@opt$convergence == 0, all_combos)

cat("Total models fitted:", length(det_model_list), "\n")
#Total models fitted: 64 

cat("Models with convergence issues removed.\n")

# QAICc model selection
step1_table <- aictab(
  cand.set = det_model_list,
  modnames = names(det_model_list),
  c.hat    = chat_det  # use your c-hat from Step 8
)


# Top models within delta QAICc <= 2
cat("\nTop occupancy models (delta QAICc <= 2):\n")
step1_table_filtered <- step1_table[step1_table$Delta_QAICc <= 2, ]
print(step1_table[step1_table$Delta_QAICc <= 2, ])



```

3 models are within ΔAICc \< 2

**Choose the simplest model with 4 covariates:**

```{r}
best_det_formula_1 <- ~visit + hunting + shifting_cultivation
best_det_formula_2 <- ~visit + area_pa + shifting_cultivation
```

```         
```

But to be sure let us do another fitcheck with these two models, a more
penalizing one.

```{r}
fitstats <- function(model.name, 
                     method = "nonparboot") {
  
  observed <- getY(model.name@data)
  expected <- fitted(model.name)
  
  resids <- residuals(model.name,
                      method = "nonparboot")
  
  sse <- sum(resids^2,
             na.rm = TRUE)
  
  chisq <- sum((observed - expected)^2 / expected,
               na.rm = TRUE)
  
  freeTuke <- sum((sqrt(observed) - sqrt(expected))^2, 
                  na.rm = TRUE)
  
  out <- c(SSE = sse,
           Chisq = chisq,
           freemanTukey = freeTuke)
  
  return(out)
  
}

pb <- parboot(det_model_list$visit_hunting_shifting_cultivation,
              fitstats,
              nsim = 1000,
              report = TRUE,
              method = "nonparboot")

pb

par(mfrow = c(3,1))

plot(pb,
     main = "",
     xlab = c("SSE", "Chisq", "FT"))

#Parametric Bootstrap Statistics:
 #               t0 mean(t0 - t_B) StdDev(t0 - t_B) Pr(t_B > t0)
#SSE           42.7           1.78             4.03        0.344
#Chisq        224.7          10.96            57.61        0.290
#freemanTukey  56.6           2.54             4.65        0.309

```

You see all the p values are \>\> 0.5 so this model(global) is a good
fit.

**Now for the next best model - visit_hunt_shift**

```{r}
pb <- parboot(det_model_list$visit_hunting.24_shifting_cultivation_evergreen_deciduous_forest,
              fitstats,
              nsim = 1000,
              report = TRUE,
              method = "nonparboot")

pb

par(mfrow = c(3,1))

plot(pb,
     main = "",
     xlab = c("SSE", "Chisq", "FT"))

```

You see all the p values are \>\> 0.5 so this model is a good fit.

Let us use
**visit_hunting.24_shifting_cultivation_evergreen_deciduous_forest**
model for modelling occupancy because it has a better QAICc score and
weight.

### Step 6 — Fit Global Occupancy Model (fix best detection)

```{r}
best_det_formula <- ~visit + hunting.24 + shifting_cultivation + evergreen_deciduous_forest # 

# Global occupancy model
global_occ <- occu(~ visit + hunting.24 + shifting_cultivation 
                   + evergreen_deciduous_forest 
                   ~ hunting.24 + shifting_cultivation + mean_elev 
                   + road_length + area_pa + evergreen_deciduous_forest
                   + canopy_height
                   + aspect + agricuture + X_populationmean.x,
                   data = umf)

summary(global_occ)
```

#### Step 7 — Simplify Global Occupancy Model

```{r}
# Extract occupancy coefficients and SEs
occ_coefs <- as.data.frame(summary(global_occ)$state)
occ_coefs

# Flag covariates where SE >= |beta|
occ_coefs$exclude <- abs(occ_coefs$SE) >= abs(occ_coefs$Estimate)
occ_coefs

# Keep covariates where exclude == FALSE
retained_occ_covs <- rownames(occ_coefs[occ_coefs$exclude == FALSE, ])
retained_occ_covs   # these go into competing models
```

Very interstingly we find that, the following have \|SE\| \> the beta
estimates:

hunting and canopy height!

indicating they do not have a significant effect on the occupancy of GH.

**Before going ahead**

#### WE note the relations of covariates on the occupancy of gh inferred by the global occupancy model

|                          |               |           |      |
|-------------------------:|---------------|-----------|------|
| **shifting_cultivation** | -2.4760111057 | 0.8497413 | =1.5 |

|               |               |           |       |
|--------------:|---------------|-----------|-------|
| **mean_elev** | -0.9065496463 | 0.4193977 | =-0.5 |

|                 |              |           |       |
|----------------:|--------------|-----------|-------|
| **road_length** | 0.7670032689 | 0.5071199 | =0.25 |

|             |              |           |      |
|------------:|--------------|-----------|------|
| **area_pa** | 0.4186929124 | 0.3182728 | =0.1 |

|                                |               |           |       |
|-------------------------------:|---------------|-----------|-------|
| **evergreen_deciduous_forest** | -1.3140488285 | 1.0400070 | =-0.3 |

|            |              |           |       |
|-----------:|--------------|-----------|-------|
| **aspect** | 0.6925081174 | 0.3610485 | =0.33 |

|                |               |           |       |
|---------------:|---------------|-----------|-------|
| **agricuture** | -1.5699133081 | 0.8178116 | =-0.7 |

|                        |               |           |       |
|-----------------------:|---------------|-----------|-------|
| **X_populationmean.x** | -0.6788002701 | 0.6241253 | =0.01 |

**Refit simplified global:**

```{r}
#  after exclusion
simplified_global_occ <- occu(~ visit + hunting.24 + shifting_cultivation 
                   + evergreen_deciduous_forest 
                   ~ shifting_cultivation + mean_elev 
                   + road_length + area_pa + evergreen_deciduous_forest
                   + aspect + agricuture + X_populationmean.x,
                   data = umf)
```

#### Step 8 — GOF on Simplified Global Occupancy Model

```{r}
gof_occ  <- mb.gof.test(simplified_global_occ, nsim = 1000, plot.hist = TRUE)
chat_occ <- gof_occ$c.hat.est
cat("c-hat for occupancy:", chat_occ, "\n")
#1.431156

```

```         
Chi-square statistic = 44.8207 
Number of bootstrap samples = 1000
P-value = 0.116

Quantiles of bootstrapped statistics:
   0%   25%   50%   75%  100% 
  8.1  21.7  27.5  35.8 362.1 

Estimate of c-hat = 1.43
```

the goodness of fit test here reveals one concern - mild overdispersion
c.hat = 1.43 (it should be close to 1). this is also evident in the
histogram of fit statistic, our chi-square statistic is 44.8 which is
some distance from the median of the simulated statistics.

two options

1.  go ahead with QAICc model selection by spefying c.hat in sicmodelavg
    function - accepted in literature
2.  Use a different detection model

**Let's try with a simpler detection model.**

```{r}
best_det_formula <- ~visit + hunting.24 + shifting_cultivation # 2nd best detection model

# Global occupancy model
global_occ_2 <- occu(~ visit + hunting.24 + shifting_cultivation 
                   ~ shifting_cultivation + mean_elev 
                   + road_length + area_pa + evergreen_deciduous_forest
                   + aspect + agricuture + X_populationmean.x,
                   data = umf)

summary(global_occ_2)

### simplify occu model
# Extract occupancy coefficients and SEs
occ_coefs <- as.data.frame(summary(global_occ_2)$state)
occ_coefs

# Flag covariates where SE >= |beta|
occ_coefs$exclude <- abs(occ_coefs$SE) >= abs(occ_coefs$Estimate)
occ_coefs

# Keep covariates where exclude == FALSE
retained_occ_covs <- rownames(occ_coefs[occ_coefs$exclude == FALSE, ])
retained_occ_covs   # these go into competing models

### refit
# Example after exclusion
simplified_global_occ <- occu(~ visit + hunting.24 + shifting_cultivation
                   ~ shifting_cultivation + mean_elev 
                   + road_length
                   + aspect + agricuture,
                   data = umf)

### gof test
gof_occ  <- mb.gof.test(simplified_global_occ, nsim = 1000, plot.hist = TRUE)
chat_occ <- gof_occ$c.hat.est
cat("c-hat for occupancy:", chat_occ, "\n")

```

```{r}
chat_occ <- gof_occ$c.hat.est
cat("c-hat for occupancy:", chat_occ, "\n")

```

```         
c-hat for occupancy: 1.284274  
```

c-hat is now closer to 1! however only by a margin, yet we loose on
important variables like human population, area under forest. So let us
use the first model only but with caution and explicitely providing
c-hat whenever assessing the further models.

#### Step 9 — Build All Competing Occupancy Models

Build singular and all additive combinations of retained covariates:

Instead of typing 128 models, use this function

```{r}

# Your retained occupancy covariates

covs <- c("shifting_cultivation", "mean_elev", "road_length", 
          "area_pa", "evergreen_deciduous_forest", "aspect", "agricuture", "X_populationmean.x")

# Detection formula
det_formula_str <- "~visit + hunting.24 + shifting_cultivation+evergreen_deciduous_forest "

# Generate all combinations
all_combos <- list()

# Null model
all_combos[[1]] <- occu(formula(paste(det_formula_str, "~1")), data = umf)
names(all_combos)[1] <- "null"

# Single covariates (7 models)
for (i in seq_along(covs)) {
  combo_name <- covs[i]
  formula_str <- paste(det_formula_str, "~", covs[i])
  all_combos[[i + 1]] <- occu(formula(formula_str), data = umf)
  names(all_combos)[i + 1] <- combo_name
}

# All combinations of 2+ covariates
for (n in 2:length(covs)) {
  combos_n <- combn(covs, n, simplify = FALSE)
  
  for (combo in combos_n) {
    combo_name <- paste(combo, collapse = "_")
    formula_str <- paste(det_formula_str, "~", paste(combo, collapse = " + "))
    
    tryCatch({
      all_combos[[length(all_combos) + 1]] <- occu(formula(formula_str), data = umf)
      names(all_combos)[length(all_combos)] <- combo_name
    }, error = function(e) {
      cat("Failed to fit:", combo_name, "\n")
    })
  }
}

# Remove models with convergence issues
occ_model_list <- Filter(function(m) m@opt$convergence == 0, all_combos)

cat("Total models fitted:", length(occ_model_list), "\n")
#Total models fitted: 503 

cat("Models with convergence issues removed.\n")

# QAICc model selection
step2_table <- aictab(
  cand.set = occ_model_list,
  modnames = names(occ_model_list),
  c.hat    = chat_occ  # use your c-hat from Step 8
)

print(step2_table)

# Top models within delta QAICc <= 2
cat("\nTop occupancy models (delta QAICc <= 2):\n")
step2_table_filtered <- step2_table[step2_table$Delta_QAICc <= 2, ]
print(step2_table_filtered)
```

#### Step 10: finding the averaged model

Now that we have the top models for occupancy let us find the average
model to find the occupancy of this species

```{r}

# Assuming your top 4 models are in a list
top_models <- list(
  model1 = occ_model_list[[73]],
  model2 = occ_model_list[[109]],
  model3 = occ_model_list[[32]],
  model4 = occ_model_list[[104]]
)

# Model-averaged occupancy parameter
## we have to use the anto logit function to extract the Backtransformed linear combination(s) of Occupancy estimate(s)

antilogit <- function(x) { exp(x) / (1 + exp(x) ) }


```

Model 1

```{r}
selected_model <- occ_model_list[[106]]
a <-summary(selected_model)
#occupancy
occu1 <- antilogit(-1.1706862)
#0.2367
det1 <- antilogit(0.7597004)
#0.6812887

```

| Col1 | Col2 | Col3 | Col4 | Col5 |
|------|------|------|------|------|
|      |      |      |      |      |
|      |      |      |      |      |
|      |      |      |      |      |
|      |      |      |      |      |
|      |      |      |      |      |
|      |      |      |      |      |

```{r}
#2
top_models[[2]]
#occupancy
occu2 <- antilogit(-1.1033502)
#0.2556014
det2 <- antilogit(0.7569210)
#0.6812887
```

```{r}
#3
top_models[[3]]
#occupancy
occu3 <- antilogit(-1.0911928)
#0.2556014
det3 <- antilogit(0.7808543)
#0.6812887
```

```{r}
#4
top_models[[4]]
#occupancy
occu4 <- antilogit(-1.1876341)
#0.2556014
det4 <- antilogit(0.7620042)
#0.6812887
```

### Now average these occupancy estimates

```{r}
avg_uccu_gh <- mean(occu1,occu2, occu3, occu4)
avg_uccu_gh
```

remember the naive was 0.3317

Greatest affectors :

distance to grid : +

shifting agriculture : -

mean elevation : -

road length : + but SE is \> beta

Agriculture : -

aspect : +

in an sub-optimal model we found

area_under_pa : +

```{r}
# road lenght
antilogit(0.3785049)
# beta = 0.5935125
antilogit(0.751)
# beta = 0.6793966

```
