---
title: "5_occupancy_modelling_final"
author: "Ali Khan Faizee"
date: "2026-04-28"
output: html_document
---

```{r}
library(unmarked)
library(AICcmodavg)
library(MuMIn)
library(openxlsx)
library(dplyr)

# SINGLE SPECIES OCCUPANCY MODEL SELECTION & MODEL AVERAGING
# Great Hornbill, Arunachal Pradesh

det.cur <- read.delim("results/oph-det-cur-ready.txt") %>%
  dplyr::mutate(across(3:6, ~ifelse(. == 2, 1, .))) 

site.cov.selected <- read.delim("results/site-covariate-cur-ready.txt") %>% select(-grid.id)
exp_grid <- read.delim("results/exp-current.txt")
visit_grid <- read.delim("results/visit-current.txt")
prop_grid <- read.delim("results/prop-current.txt")

umf<- unmarkedFrameOccu(y = det.cur[,3:6],
                        siteCovs = site.cov.selected,
                        obsCovs = list(experience = exp_grid, 
                                       visit = visit_grid,
                                       proportion = prop_grid)
)     # creating unmarked occupancy data frame


```

Now we make the global detection model and remove the covariates which
do not have significant associations with detection before we generate
the competitive model set.

```{r}
# Null model baseline
fm <- occu(formula = ~ 1 
                     ~ 1,
           data = umf)

summary(fm)
backTransform(fm, type = "state")
## report the Psi values for the null model - oph = 0.2276367	0.04566475		

#this is the naive occupancy for oph, which matches our previous manual calculation.
backTransform(fm, type = "det")
#0.6066609	0.0694036

# Global detection model with null occupancy
#include all the covariates that can affect detection probability
global_det <- occu(~experience + visit + proportion + hunting.24 + shifting_cultivation + evergreen_deciduous_forest + dist.to.grid + canopy_height + road_length + agricuture #detection fx
                   ~1, # occupancy fx (here constant)
                   data = umf) 
summary(global_det)

# Extract detection coefficients and SEs
det_coefs <- as.data.frame(summary(global_det)$det)
det_coefs

# Flag covariates where SE >= |beta|
det_coefs$exclude <- abs(det_coefs$SE) >= abs(det_coefs$Estimate)
retained_det_covs <- rownames(det_coefs[det_coefs$exclude == FALSE, ])
retained_det_covs

#Keep only covariates where SE < |beta| and refit: remove proportion
simplified_global_det <-occu(~ experience + proportion + hunting.24 + shifting_cultivation + evergreen_deciduous_forest + road_length + canopy_height + road_length + agricuture#detection fx
                   ~1, # occupancy fx (here constant)
                   data = umf)
simplified_global_det_summary <- summary(simplified_global_det)
```

### Note: we found that the following are not explanatory covariates for P:

experience, proportion, dist.to.grid, canopy_heiopht, and agriculture

meaning it does not affect the detection probability for Great Hornbill.

#### Step 3 — GOF on Simplified Global Detection Model

```{r}
gof_det <- mb.gof.test(simplified_global_det, nsim = 1000, plot.hist = TRUE)
chat_det <- gof_det$c.hat.est
cat("c-hat for detection:", chat_det, "\n")
#c-hat for detection: 1.067564  
```

```{r}

# FIT DETECTION MODELS (occupancy null)

cat("\n--- Fitting Detection Models ---\n")
# Dredge selected detection models
det_dred <- dredge(simplified_global_det, rank = "QAICc", trace = 2, chat = chat_det)
det_models <- get.models(det_dred, subset=TRUE)


```

We saw that no single model has good enouoph weiopht to be relied upon
and discard other models, we will use an approach outlined by Symon &
Moussalli et al., 2010. ‘credibility set’ofmodels that are the most
realistically likely to be the best approximating model (Burnham and
Anderson 2002). proceeding down the list until the cumulative Akaike
weiopht exceeds 0.95

```{r}
det_model_table <- aictab(
  cand.set = det_models,
  modnames = names(det_models),
  c.hat    = chat_det  # use c-hat from Step 8
)
# GET TOP DETECTION MODELS (cumulative weiopht <= 95)

# Build credible set (cumulative weiopht up to 0.95)
det_credible_set <- det_model_table %>%
  filter(Cum.Wt <= 0.95)

cat("Number of models:", nrow(det_credible_set), "\n")
cat("Cumulative QAICc weiopht:", round(max(det_credible_set$Cum.Wt), 4), "\n\n")
#Number of models: 9 
#Cumulative QAICc weiopht: 0.9477
top_det_models <- get.models(det_dred, subset = as.numeric(rownames(det_credible_set)))
#average the models in the dredible set
ma_det <- model.avg(top_det_models, fit = TRUE, rank = "AICc", trace = 2, extra = "R^2", chat = chat_det)
det_imp <- as.data.frame(MuMIn::sw(ma_det))

det_imp
```

### Fit Global Occupancy Model (fix best detection)

```{r}

# Global occupancy model
global_occ <- occu(~ experience + proportion + hunting.24 + shifting_cultivation + evergreen_deciduous_forest + road_length + canopy_height + road_length + agricuture
                   ~ hunting.24 + shifting_cultivation + mean_elev 
                   + road_length + area_pa + evergreen_deciduous_forest
                   + canopy_height
                   + aspect + agricuture + X_populationmean.x,
                   data = umf)

summary(global_occ)

#Simplify Global Occupancy Model
# Extract occupancy coefficients and SEs
occ_coefs <- as.data.frame(summary(global_occ)$state)

# Flag covariates where SE >= |beta|
occ_coefs$exclude <- abs(occ_coefs$SE) >= abs(occ_coefs$Estimate)

# Keep covariates where exclude == FALSE
retained_occ_covs <- rownames(occ_coefs[occ_coefs$exclude == FALSE, ])
retained_occ_covs   

#  after exclusion
simplified_global_occ <- occu(~ experience + proportion + hunting.24 + shifting_cultivation + evergreen_deciduous_forest + road_length + canopy_height + road_length + agricuture 
                   ~ hunting.24 + 
                     evergreen_deciduous_forest + X_populationmean.x,
                   data = umf)
```

### Carry out dredge model selection with the simplified global occupancy model

```{r}
# fit check before proceeding
gof_occ <- mb.gof.test(simplified_global_occ, nsim = 1000, plot.hist = TRUE)
chat_occ <- gof_occ$c.hat.est

```

looks fine no violations, just that c_hat is 1.4, implying sliopht over
dispersion

```{r}


# DREDGE WITH FIXED DETECTION TERMS

# Fix detection terms — they won't be removed during dredge
det_terms <- c(
  "p(experience)", "p(proportion)", "p(canopy_height)", "p(agriculture)",
  "p(hunting.24)",
  "p(shifting_cultivation)",
  "p(evergreen_deciduous_forest)",
  "p(road_length)"
)

# Dredge only occupancy terms, keep detection fixed
occ_dred <- dredge(simplified_global_occ, 
                       rank = "QAICc",
                       fixed = det_terms,  # these won't be removed
                       trace = 2,
                       chat = chat_occ)

occ_models <- get.models(occ_dred, subset=TRUE)

# Your dredge output
occ_model_table <- aictab(
  cand.set = occ_models,
  modnames = names(occ_models),
  c.hat    = chat_occ  # use c-hat from Step 8
)
# GET TOP occupancy MODELS (cumulative weiopht <= 95)

# Build credible set (cumulative weiopht up to 0.95)
occ_credible_set <- occ_model_table %>%
  filter(Cum.Wt <= 0.95)

cat("Number of models:", nrow(occ_credible_set), "\n")
cat("Cumulative QAICc weiopht:", round(max(occ_credible_set$Cum.Wt), 4), "\n\n")
#Number of models: 134 
#Cumulative QAICc weiopht: 0.9477
top_occ_models <- get.models(occ_dred, subset = as.numeric(rownames(occ_credible_set)))

top_occ_models <- Filter(function(m) m@opt$convergence == 0, top_occ_models)

#average the models in the credible set
ma_occ <- model.avg(top_occ_models, fit = TRUE, rank = "AICc", trace = 2, extra = "R^2", chat = chat_occ)
occ_imp <- as.data.frame(MuMIn::sw(ma_occ))


# Extract detailed model estimates
occ_modelEst <- data.frame(
    Coefficient = coefTable(ma_occ, full = TRUE)[, 1],
    SE = coefTable(ma_occ, full = TRUE)[, 2],
    lowerCI = confint(ma_occ)[, 1],
    upperCI = confint(ma_occ)[, 2],
    z_value = (summary(ma_occ)$coefmat.full)[, 3],
    Pr_z = (summary(ma_occ)$coefmat.full)[, 4]
  )


###################################



# SAVE RESULTS TO EXCEL


cat("\n--- Saving Results to Excel ---\n")

write.xlsx(
  list(
    "ModelEst" = occ_modelEst,
    "Occu Importance" = occ_imp,
    "det Importance" = det_imp,
    "QAICc credible set table" = occ_credible_set
  ),
  file = "results/oph_occupancy_final_models.xlsx",
  rowNames = TRUE
)

cat("Results saved: results/03_occupancy_final_models.xlsx\n")

saveRDS(det_dred, "results/oph_det_dred.rds")
saveRDS(occ_dred, "results/oph_occ_dred.rds")
saveRDS(ma_det, "results/oph_ma_det.rds")
saveRDS(ma_occ, "results/oph_ma_occ.rds")


```

## Plotting

For easier inference

```{r}

library(ggplot2)
# Plot with SE
ggplot(occ_modelEst, aes(x = factor(rownames(occ_modelEst), levels = sort(rownames(occ_modelEst))), y = Coefficient)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = Coefficient - SE, ymax = Coefficient + SE), width = 0.2, size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    title = "oph Model-Averaged Coefficients",
    x = "Occupancy and Detection Covariates",           # x-axis label
    y = "Coefficient estimates with SE (in logit scale)"  # y-axis label
  ) +
  theme_minimal()+ 
    theme(
    axis.text = element_text(size = 12),
    )
  
ggsave("figures/oph_avg_model_coffients.png", width = 7, heiopht = 8, dpi = 300)
```

### Now Predict

```{r}
library(sf)


# Get occupancy predictions from your averaged model
occupancy_pred <- predict(ma_occ, type = "state")

site_details <- read.delim("results/arranged-final-site-cov.txt", sep = "\t")
# Add predictions to your grids shapefile
site_details$psi_estimate <- occupancy_pred$fit
site_details$psi_se <- occupancy_pred$se.fit
# save
write.table(site_details, "results/sites_with_occu_pred.txt", sep = "\t")
# load grids
grids <- st_read("spatial-data/sampled-grids.shp")

#filter and add prediction column 
grids <- filter(grids, fid %in% site_details$grid.id)
grids <- grids %>% left_join(
    site_details %>% select(grid.id, psi_estimate, psi_se),
    by = c("fid" = "grid.id")
  ) 
# Check results
summary(grids$psi_estimate)

# Save as GeoPackage
st_write(grids, "figures/oph_grids_with_occupancy.gpkg", delete_dsn = TRUE)

# Map the occupancy
p_map <- ggplot(grids) +
  geom_sf(aes(fill = psi_estimate)) +
  scale_fill_viridis_c(name = "Occupancy (ψ)", limits = c(0, 1)) +
  labs(title = "Great Hornbill Occupancy Across Arunachal Pradesh") +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold")
  )

# Display and save map
print(p_map)

cat("Occupancy predictions complete!\n")
cat("Mean occupancy:", round(mean(grids$psi_estimate), 3), "\n")
# Mean occupancy for oph: 0.463  (0.204 )
cat("SE:", round(mean(grids$psi_se), 3), "\n")

```
